A sample workflow

### 2025-04-24 analyses/african_mammals_workflow.qmd
some stuff I worked on, on the plane, still pretty much a draft

---
title: "A sample workflow"
authors: "Lizzie, Victor"
date: "April 2025"
date-format: "MMMM YYYY"
toc: true
number-sections: true
highlight: pygments
execute:
  enabled: true
format:
  html:
    html-math-method: katex
    theme:
      - lux
    embed-resources: true
    code-overflow: wrap
    linkcolor: "#B97C7C"
    page-layout: full
    fig-width: 8
    fig-height: 4
  pdf:
    keep-tex: true
    fig-width: 5.5
    fig-height: 5.5
    code-overflow: wrap
    monofontoptions:
      - Scale=0.5
knitr:
  opts_chunk:
    comment: ''
  opts_knit:
    global.par: TRUE
format-links: false
---

@@

```{r}
set.seed(20012029)

@@

@@

### Simulating data
Now we start to simulate!
*Sorry, I did not have time to do this, but I would simulate from the above model or some other simple one you come up with. I would simulate really NICE data, not realistic data.* 
```{r}

nspecies <- 9
npops_persp <- round(runif(nspecies, 1, 3)) # no. of different populations per species
npops <- sum(npops_persp) # total no. of different populations
nobs_perpop <- round(runif(npops, 5, 15)) # no. of different observations per population

@@

years <- c()
for(np in nobs_perpop){
  years_p <- sample(1960:2020, size = np, replace = FALSE)
  years <- c(years, years_p)
}

# nested intercepts
mu_species_alpha <- rnorm(n = nspecies, mean = 200, sd = 25)
sig_species_alpha <- rnorm(n = nspecies, mean = 5, sd = 2)
mu_pop_alpha <- rnorm(n = npops, mean = mu_species_alpha[spid_perpop])
sig_pop_alpha <- abs(rnorm(n = npops))
alpha <- rnorm(n = npops, mean = mu_pop_alpha, sd = sig_pop_alpha)

# nested slopes
mu_species_beta <- rnorm(n = nspecies, mean = 0, sd = 0.1)
sig_species_beta <- rnorm(n = nspecies, mean =0, sd = 0.5)
mu_pop_beta <- rnorm(n = npops, mean = mu_species_beta[spid_perpop], sd = 0.2)
sig_pop_beta <- abs(rnorm(n = npops))
beta <- rnorm(n = npops, mean = mu_pop_beta, sd = sig_pop_beta)

sigma_obs <- abs(rnorm(1, mean = 10, sd = 1))

@@

  yhat_p <- alpha_p + beta_p * (years_p - 1980)
  # yhat_p <- alpha_p
  y_p <- rnorm(n = nobs_p, mean = yhat_p, sd = sigma_obs)
  y <- c(y, y_p)
}

dataplot <- data.frame(
  y,
  year = years - 1980,
  species = as.character(spid_perobs),
  population = as.character(popid_perobs)
)

@@

nestlm <- lm(y ~ year * species/population, data = dataplot)
coefs <- summary(nestlm)$coefficients

species_intercept <- c(coefs["(Intercept)", "Estimate"], sapply(2:nspecies, function(sp){coefs["(Intercept)", "Estimate"] + coefs[paste0("species", sp), "Estimate"]}))
species_intercept_sd <- c(coefs["(Intercept)", "Std. Error"], sapply(2:nspecies, function(sp){coefs[paste0("species", sp), "Std. Error"]}))

species_slope <- c(coefs["year", "Estimate"], sapply(2:nspecies, function(sp){coefs["year", "Estimate"] + coefs[paste0("year:species", sp), "Estimate"]}))
species_slope_sd <- c(coefs["year", "Std. Error"], sapply(2:nspecies, function(sp){coefs[paste0("year:species", sp), "Std. Error"]}))

@@

slopes <- data.frame(obs=mu_species_beta, est=species_slope, est.sd=species_slope_sd, species = as.character(1:nspecies))
ggplot(data = slopes) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey") + 
  geom_pointrange(aes(ymin = est-est.sd, ymax = est+est.sd, x = obs, y = est, color = species)) +
  theme_bw()

```
*Add some text here showing the model output returning parameters.*

@@

```{r}

@@

@@

```{r}

lpi_subset_rsh$year <- lpi_subset_rsh$year-1980
nestlm <- lm(value ~ year * genus_species/ID, data = lpi_subset_rsh)
coefs <- summary(nestlm)$coefficients

@@

```{r}
#| fig-height: 10
#| fig-width: 12

retrodata <- data.frame(obs = lpi_subset_rsh$value, pred = predict(nestlm),
                        species = lpi_subset_rsh$genus_species, id = lpi_subset_rsh$ID)

@@

>>>>

---
title: "A sample workflow"
authors: "Lizzie, Victor"
date: "April 2025"
date-format: "MMMM YYYY"
toc: true
number-sections: true
highlight: pygments
execute:
  enabled: true
format:
  html:
    html-math-method: katex
    theme:
      - lux
    embed-resources: true
    code-overflow: wrap
    linkcolor: "#B97C7C"
    page-layout: full
    fig-width: 8
    fig-height: 4
  pdf:
    keep-tex: true
    fig-width: 5.5
    fig-height: 7.5
    code-overflow: wrap
    monofontoptions:
      - Scale=0.5
knitr:
  opts_chunk:
    comment: ''
  opts_knit:
    global.par: TRUE
format-links: false
---

@@

```{r}
wd <- "/home/victor/projects/forecastflows/analyses"

set.seed(20012029)

@@

lpi_subset <- read.table(file.path(wd, "jdavies/African_mammals.txt"))

@@

### Simulating data
Now we start to simulate!
*Sorry, I did not have time to do this, but I would simulate from the above model or some other simple one you come up with. I would simulate really NICE data, not realistic data.* 
```{r sim data}

nspecies <- 9
npops_persp <- round(runif(nspecies, 2, 15)) # no. of different populations per species
npops <- sum(npops_persp) # total no. of different populations
nobs_perpop <- round(runif(npops, 10, 40)) # no. of different observations per population

@@

years <- c()
for(np in nobs_perpop){
  years_p <- sample(1950:2020, size = np, replace = FALSE)
  years <- c(years, years_p)
}

# nested intercepts
mu_species_alpha <- rnorm(n = nspecies, mean = log(200), sd = log(5))
sig_species_alpha <- rnorm(n = nspecies, mean = 5, sd = 5)
mu_pop_alpha <- rnorm(n = npops, mean = mu_species_alpha[spid_perpop])
sig_pop_alpha <- abs(rnorm(n = npops))
alpha <- rnorm(n = npops, mean = mu_pop_alpha, sd = sig_pop_alpha)

# nested slopes
mu_species_beta <- rnorm(n = nspecies, mean = 0, sd = 0.01)
sig_species_beta <- rnorm(n = nspecies, mean = 0, sd = 0.05)
mu_pop_beta <- rnorm(n = npops, mean = mu_species_beta[spid_perpop], sd = 0.05)
sig_pop_beta <- abs(rnorm(n = npops, mean = 0, sd = 0.01))
beta <- rnorm(n = npops, mean = mu_pop_beta, sd = sig_pop_beta)

sigma_obs <- abs(rnorm(1, mean = 0.25, sd = 0.25))

@@

  yhat_p <- alpha_p + beta_p * (years_p - 1980)
  # yhat_p <- alpha_p
  y_p <- exp(rnorm(n = nobs_p, mean = yhat_p, sd = sigma_obs))
  y <- c(y, y_p)
}

dataplot <- data.frame(
  y, 
  year = years - 1980,
  species = as.character(spid_perobs),
  population = paste0(spid_perobs,popid_perobs)
)

dataplot$logy <- log(dataplot$y)

@@

ggplot(data = dataplot) +
  geom_line(aes(x = year+1980, y = logy, 
                group = paste0(species, population),
                color = species)) +
  theme_bw()

### Fit model on simulated data!

```{r fit sim data}

nestlm <- lm(logy ~ species/population + year * species/population, data = dataplot)
coefs <- summary(nestlm)$coefficients

#### Retrodictive check with model fitted on simulated data

```{r retro sim data}
#| fig-height: 6
#| fig-width: 8

retrodata <- data.frame(obs = dataplot$logy, pred = predict(nestlm),
                        year = dataplot$year,
                        species = dataplot$species, id = dataplot$population)

ggplot(data = retrodata) +
  facet_wrap(~ paste0("Species ", species), scales = "free") +
  geom_point(aes(x = year, y = obs, group = id, color = as.character(id)),
             size = 1) + 
  geom_line(aes(x = year, y = pred, group = id, color = as.character(id))) +
  theme_bw() + theme(legend.position = 'none')

ggplot(data = retrodata) +
  facet_wrap(~ paste0("Species ", species), scales = "free") +
  geom_point(aes(x = year, y = exp(obs), group = id, color = as.character(id)),
             size = 1) + 
  geom_line(aes(x = year, y = exp(pred), group = id, color = as.character(id))) +
  theme_bw() + theme(legend.position = 'none')

#### Inference with model fitted on simulated data

The model does a poor job at recovering species-level parameters.

```{r inference sim data}
species_intercept <- c(coefs["(Intercept)", "Estimate"], sapply(2:nspecies, function(sp){coefs["(Intercept)", "Estimate"] + coefs[paste0("species", sp), "Estimate"]}))
species_intercept_sd <- c(coefs["(Intercept)", "Std. Error"], sapply(2:nspecies, function(sp){coefs[paste0("species", sp), "Std. Error"]}))

species_slope <- c(coefs["year", "Estimate"], sapply(2:nspecies, function(sp){coefs["year", "Estimate"] + coefs[paste0("species", sp, ":year"), "Estimate"]}))
species_slope_sd <- c(coefs["year", "Std. Error"], sapply(2:nspecies, function(sp){coefs[paste0("species", sp,":year"), "Std. Error"]}))

@@

slopes <- data.frame(obs=mu_species_beta, est=species_slope, est.sd=species_slope_sd, species = as.character(1:nspecies))
ggplot(data = slopes) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey") + 
  geom_pointrange(aes(ymin = est-est.sd, ymax = est+est.sd, x = obs, y = est, color = species)) +
  theme_bw()
```

@@

```{r fitdata}
#| eval: true

@@

ggplot(data = lpi_subset_rsh[lpi_subset_rsh$value != 0,]) +
  geom_line(aes(x = year, y = log(value), 
                group = paste0(genus_species, ID),
                color = genus_species)) +
  theme_bw()

@@

```{r}
#| eval: false

runmodel <- TRUE

lpi_subset_rsh <- lpi_subset_rsh[lpi_subset_rsh$value != 0,]
lpi_subset_rsh$year <- lpi_subset_rsh$year-1980
lpi_subset_rsh$logvalue <- log(lpi_subset_rsh$value)
lpi_subset_rsh$ID <- as.character(lpi_subset_rsh$ID)

if(runmodel){
  nestlm <- lm(logvalue ~ genus_species/ID + year * genus_species/ID, data = lpi_subset_rsh)

  saveRDS(nestlm, file = file.path(wd, "output/fit/nestmodel1.rds"))
}else{
  nestlm <- readRDS(file.path(wd, "output/fit/nestmodel1.rds"))
}

coefs <- summary(nestlm)$coefficients

@@

```{r}
#| fig-height: 10
#| fig-width: 12
#| eval: false

retrodata <- data.frame(obs = lpi_subset_rsh$logvalue, pred = predict(nestlm),
                        year = lpi_subset_rsh$year, units = lpi_subset_rsh$Units,
                        species = lpi_subset_rsh$genus_species, id = lpi_subset_rsh$ID)

@@

```{r}
#| fig-height: 8
#| fig-width: 8
#| eval: false

ggplot(data = retrodata[retrodata$species == "Loxodonta africana", ]) +
  facet_wrap(~ id, scales = "free") +
  geom_point(aes(x = year, y = obs, group = id, color = id),
             size = 1) + 
  geom_line(aes(x = year, y = pred, group = id, color = id)) +
  theme_bw() + theme(legend.position = 'none')
